A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s

### 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex

Over the past 20 years the HEP community has developed and gravitated
around a single analysis ecosystem based on ROOT. 
ROOT is a general-purpose object
oriented framework that addresses the selection, integration,
development, and support of a number of foundation and utility class
libraries that can be used as a basis for developing HEP application
codes. The added value to the HEP community is that it provides an
integrated and validated toolkit, and its use encompasses the full event
processing chain; it has a major impact on the way HEP analysis is
performed. This lowers the hurdle to start an analysis, enabling the
community to communicate using a common analysis language, as well as
making common improvements as additions to the toolkit quickly become
available. The ongoing ROOT programme of work addresses important new
requirements, in both functionality and performance, and this is given
a high priority by the HEP community.

An important new development in the analysis domain has been the
emergence of new analysis tools coming from industry and open source
projects, and this presents new opportunities for improving the HEP
analysis software ecosystem. The HEP community is very interested in
using these software tools, together with established components, in an
interchangeable way. The main challenge will be to enable new
open-source tools to be plugged in dynamically to the existing ecosystem
and to provide mechanisms that allow the existing and new components to
interact and exchange data efficiently. To improve our ability to
analyse much larger datasets, R\&D will be needed to investigate file
formats, compression algorithms, and new ways of storing and accessing
data for analysis and to adapt workflows to run on future computing
infrastructures.

Reproducibility is the cornerstone of scientific results. It is
currently difficult to repeat most HEP analyses in exactly the manner
they were originally performed. This difficulty mainly arises due to the
number of scientists involved, the large number of steps in a typical
HEP analysis workflow, and the complexity of the analyses themselves. A
challenge specific to data analysis and interpretation is tracking the
evolution of relationships between all the different components of an
analysis.

@@

One new model of data analysis, developed outside of HEP, maintains the
concept of sequential reduction, but mixes interactivity with
batch processing. These exploit new cluster management systems, most
notably Apache Spark~\cite{Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
industry and the data-science community. Other products implementing the
same analysis concepts and workflows are emerging, such as TensorFlow~\cite{tensorflow2015-whitepaper},
Dask~\cite{dask_ref}, Pachyderm~\cite{thepachydermteam}, 
Blaze~\cite{blaze}, Parsl~\cite{babuji_yadu_2017_853492}, 
and Thrill~\cite{bingmann2016thrill}. This approach can complement
the present and widely adopted Grid processing of datasets. It may
potentially simplify the access to data and the expression of
parallelism, thereby improving the exploitation of cluster resources.

>>>>

Over the past 20 years the HEP community has developed and gravitated
around a single analysis ecosystem based on ROOT.

ROOT is a general-purpose object
oriented framework that addresses the selection, integration,
development, and support of a number of foundation and utility class
libraries that can be used as a basis for developing HEP application
codes. The added value to the HEP community is that it provides an
integrated and validated toolkit, and its use encompasses the full event
processing chain; it has a major impact on the way HEP analysis is
performed. This lowers the hurdle to start an analysis, enabling the
community to communicate using a common analysis language, as well as
making common improvements as additions to the toolkit quickly become
available. The ongoing ROOT programme of work addresses important new
requirements, in both functionality and performance, and this is given
a high priority by the HEP community.

An important new development in the analysis domain has been the
emergence of new analysis tools coming from industry and open-source
projects (e.g. Jupyter notebooks~\cite{http://jupyter.org/}, the scikit-learn package~\cite{scikit-learn}), and this presents new opportunities for improving the HEP
analysis software ecosystem. The HEP community is very interested in
using these software tools, together with established components, in an
interchangeable way. The main challenge will be to enable new
open-source tools to be plugged in dynamically to the existing ecosystem
and to provide mechanisms that allow the existing and new components to
interact and exchange data efficiently. To improve our ability to
analyse much larger datasets, R\&D will be needed to investigate file
formats, compression algorithms, and new ways of storing and accessing
data for analysis and to adapt workflows to run on future computing
infrastructures.

Reproducibility is the cornerstone of scientific results. It is
currently difficult to repeat most HEP analyses in exactly the manner
they were originally performed. This difficulty mainly arises due to the
number of scientists involved, the large number of steps in a typical
HEP analysis workflow, and the complexity of the analyses themselves. A
challenge specific to data analysis and interpretation is tracking the
evolution of relationships between all the different components of an
analysis, i.e. the provenance of each step.

Reproducibility of scientific results goes in hand with the need to preserve both the data and the software.
Section~\ref{data-and-software-preservation}} develops this latter topic where the FAIR principles of data management are embraced.

@@

One new model of data analysis, developed outside of HEP, maintains the
concept of sequential reduction, but mixes interactivity with
batch processing. These exploit new cluster management systems, most

notably Apache Spark~\cite{Spark,Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
industry and the data-science community. Other products implementing the
same analysis concepts and workflows are emerging, such as TensorFlow~\cite{tensorflow2015-whitepaper},
Dask~\cite{rocklin2015dask,dask_ref}, Pachyderm~\cite{thepachydermteam}, 
Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, 
and Thrill~\cite{bingmann2016thrill}. This approach can complement
the present and widely adopted Grid processing of datasets. It may

potentially simplify the access to data and the expression of
parallelism, thereby improving the exploitation of cluster resources.
